---
title: "6306 - Case Study 1"
author: "Josh"
date: "2026-03-5"
output: html_document
editor_options: 
  chunk_output_type: console
---

### LINK TO PRESENTATION
https://youtu.be/riuNzlxg1Sw

### Libraries
```{r}
library(dplyr)
library(tidyverse)
library(class)
library(caret)
library(e1071)
library(ggplot2)
library(GGally)
library(scales)
```

### Reading in Main Dataset
```{r}
df <- read.csv('/Users/joshtsamuel/SMU/DS-6306 Doing Data Science/Units/Unit 8 & 9 - Case Study/CaseStudy1-data.csv')
```

### Checking for NAs
```{r}
library(naniar)
gg_miss_var(df)
# No NAs
```

### Data Wrangling
```{r}
df <- df %>%
  mutate(
    BusinessTravel = ordered(BusinessTravel,
                        levels = c("Non-Travel", "Travel_Frequently", "Travel_Rarely")),
    Department = ordered(Department,
                          levels = c("Human Resources", "Research & Development", "Sales")),
    EducationField = ordered(EducationField,
                             levels = c ("Human Resources", "Life Sciences", "Marketing", "Medical", "Other", "Technical Degree")),
  )
```

### Adding Annual_Salary Column
```{r}
df = df %>% mutate(Annual_Salary = 12*MonthlyIncome)
# getting average salary of employee over all observations
avg_sal = mean(df$Annual_Salary)
avg_sal
```

## Start of EDA

### GGPairs
```{r}
# Note: # A - Attrition (Response Variable) : Yes or No

df %>% select(DistanceFromHome, YearsAtCompany, WorkLifeBalance, YearsSinceLastPromotion, Attrition) %>% ggpairs()
# medians for DistanceFromHome for A-Y and A-N look very similar but the spread for Y is a lot larger
# medians for YearsatCompany is higher for A-N than A-Y
# most common WorkLifeBalance score is 3
# median & overall distribution HourlyRate is slightly greater for A-Y than for A-N. I would have expected the opposite

df %>% select(DailyRate, EducationField, EnvironmentSatisfaction, YearsSinceLastPromotion, Attrition) %>% ggpairs()
# median for YearsSinceLastPromotion are very similar for A-N and A-Y. I would have expected a greater difference
# median for DailyRate is  slightly greater for A-N than for A-Y.
# Each of the EducationFields had similiar distributions

df %>% select(OverTime, PercentSalaryHike, TotalWorkingYears, YearsInCurrentRole, Attrition) %>% ggpairs()
# overall distribution for YearsInCurrentRole appears greater for A-N than A-Y however medians are close. Wonder if difference is statistically significant. 
# PercentSalaryHike appears nearly identical for both  A-Y and A-N
# 0.487 correlation between YearsInCurrentRole & TotalWorkingYears which makes sense

df %>% select(DailyRate, HourlyRate, MonthlyRate, MonthlyIncome, Attrition) %>% ggpairs()
# MonthlyIncome median and overall distribution is greater for A-N than A-Y which is expected
# Strangely enough, there appears to be very little correlation between MonthlyIncome, HourlyRate, MonthlyRate, and DailyRate. I would have expected to see that as one increases, the other increases however that is not the case.

df %>% select(BusinessTravel, Age, MonthlyIncome, Attrition) %>% ggpairs()
# 0.484 correlation between Age & MonthlyIncome
# Median Age and overall distribution appears greater for A-N than A-Y which I think makes intuitive sense due to younger employees typically being more inclined to look for better opportunities. 

df %>% select(JobLevel, PerformanceRating, MonthlyIncome, YearsWithCurrManager, Attrition) %>% ggpairs()
# 0.952 correlation between MonthlyIncome and JobLevel which makes intuitive sense
# 0.367 correlation between YearsWithCurrManager and JobLevel which also makes intuitive sense
# 0.328 correlation between YearsWithCurrManager and MonthlyIncome
# YearswithCurrManager appears to have slightly greater median and distribution for A-N than A-Y

df %>% select(MaritalStatus, Department, TrainingTimesLastYear, NumCompaniesWorked,Gender, Attrition) %>% ggpairs()
# for A-N, married is highest %. For A-Y, single is highest %
```


### Smaller GGPairs based on results
```{r}
df %>% select(MonthlyIncome, JobLevel) %>% ggpairs()
df %>% select(Age, MonthlyIncome) %>% ggpairs()
df %>% select(Age, Attrition) %>% ggpairs()
df %>% select(MonthlyIncome, Attrition) %>% ggpairs()
df %>% select(JobLevel, Attrition) %>% ggpairs()

library(ggplot2)
library(scales)

df2 <- df %>% mutate(x = factor(MaritalStatus),y = factor(Attrition))  # keep as "No/Yes" or "0/1"

ggplot(df2, aes(x = x, fill = y)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Mairtal Status", y = "Attrition Percentage", fill = "Attrition", title = "Attrition Percentage by Marital Status") +
  theme_minimal()
```

```{r}
# Percentage of Yes vs No for Attrition
df %>%
  count(Attrition) %>%
  mutate(Percent = n / sum(n) * 100)

# when Age < 30
df %>%
  filter(Age < 30) %>%
  summarise(
    Percent_Yes = mean(Attrition == "Yes") * 100
  )

# when BusinessTravel = Travel_Frequently
df %>%
  filter(BusinessTravel == "Travel_Frequently") %>%
  summarise(
    Percent_Yes = mean(Attrition == "Yes") * 100
  )

# histogram of years since last promotion
hist(df$YearsSinceLastPromotion)

# when YearsSinceLastPromotion is more than 8 years
df %>%
  filter(YearsSinceLastPromotion > 8) %>%
  summarise(
    Percent_Yes = mean(Attrition == "Yes") * 100
  )
```


### Subsetting Dataset - df1 (only removing filler variables)
```{r}
df1 = df %>%
  select(-EmployeeCount, -Over18, -StandardHours)
```


### Looking at Factors with biggest contribution to attrition
```{r}
# Monthly Income
df1 %>% ggplot(aes(x = Attrition, y = MonthlyIncome, fill = Attrition)) + geom_boxplot() + labs(title = "Monthly Income by Attrition Status", x = "Attrition", y = "Monthly Income")  + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# OverTime
ot_summary <- df1 %>%
  group_by(OverTime) %>%
  summarise(AttritionRate = mean(Attrition == "Yes")) 
ggplot(ot_summary, aes(x = OverTime, y = AttritionRate, fill = OverTime)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Attrition Rate by Overtime Status",
       x = "Overtime",
       y = "Attrition Rate") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# JobLevel
jl_summary <- df1 %>%
  group_by(JobLevel) %>%
  summarise(AttritionRate = mean(Attrition == "Yes"))
ggplot(jl_summary, aes(x = JobLevel, y = AttritionRate, fill = JobLevel)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Attrition Rate by Job Level",
       x = "Job Level",
       y = "Attrition Rate") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# EducationField
ef_summary <- df1 %>%
  group_by(EducationField) %>%
  summarise(AttritionRate = mean(Attrition == "Yes"))
ggplot(ef_summary, aes(x = EducationField, y = AttritionRate, fill = EducationField)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Attrition Rate by Education Field",
       x = "Education Field",
       y = "Attrition Rate") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# Marital Status
ms_summary <- df1 %>%
  group_by(MaritalStatus) %>%
  summarise(AttritionRate = mean(Attrition == "Yes"))
ggplot(ms_summary, aes(x = MaritalStatus, y = AttritionRate, fill = MaritalStatus)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Attrition Rate by Marital Status",
       x = "Marital Status",
       y = "Attrition Rate") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# Age
df1 %>% ggplot(aes(x = Attrition, y = Age, fill = Attrition)) + geom_boxplot() + labs(title = "Age by Attrition Status", x = "Attrition", y = "Age") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
# TotalWorkingYears
df1 %>% ggplot(aes(x = Attrition, y = TotalWorkingYears, fill = Attrition)) + geom_boxplot() + labs(title = "Total Working Years by Attrition Status", x = "Attrition", y = "Total Working Years") + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 13), legend.position = "none"
)
```

## Modeling

### NB - including all metrics besides filler variables
```{r}
set.seed(6)

iterations = 100
masterAcc  = numeric(iterations)
masterSens = numeric(iterations)
masterSpec = numeric(iterations)

splitPerc = .7
threshold = 0.30   # <---- change this to experiment
# makes decision threshold for predicting "Yes" to be .30 instead of traditional ">.50"


for(j in 1:iterations)
{
  trainIndices = sample(1:nrow(df1), round(splitPerc * nrow(df1)))
  train = df1[trainIndices,]
  test  = df1[-trainIndices,]
  
  model = naiveBayes(train[,c(2, 4:33)], as.factor(train$Attrition), laplace = 1)
  
  # Get predicted probabilities
  probs = predict(model, test[,c(2, 4:33)], type = "raw")
  
  # Apply custom threshold
  pred = ifelse(probs[, "Yes"] > threshold, "Yes", "No")
  pred = as.factor(pred)
  
  CM = confusionMatrix(pred, as.factor(test$Attrition), positive = "Yes")
  
  masterAcc[j]  = CM$overall["Accuracy"]
  masterSens[j] = CM$byClass["Sensitivity"]
  masterSpec[j] = CM$byClass["Specificity"]
}

MeanAcc  = mean(masterAcc)
MeanSens = mean(masterSens)
MeanSpec = mean(masterSpec)
```

### Seeing which variables are the strongest predictors
```{r}
model <- naiveBayes(Attrition ~ .- ID, data = train)
model$tables
```

### Subsetting Dataset - df1A
```{r}
df1A = df %>% select(ID, Attrition, MonthlyIncome, TotalWorkingYears, OverTime, JobLevel, YearsAtCompany, YearsInCurrentRole, NumCompaniesWorked, EducationField, StockOptionLevel, Age, MaritalStatus, Department, JobRole, JobInvolvement, JobSatisfaction)
```

### NB - looking at  Accuracy, Sensitivity, Specificty - df1A
```{r}
set.seed(6)

iterations = 100
# getting averages for acc, sens, spec
masterAcc  = numeric(iterations)
masterSens = numeric(iterations)
masterSpec = numeric(iterations)

# getting averages for confusion matrix
masterTP = numeric(iterations)
masterTN = numeric(iterations)
masterFP = numeric(iterations)
masterFN = numeric(iterations)

splitPerc = .7
threshold = 0.30   # <---- change this to experiment
# makes decision threshold for predicting "Yes" to be .30 instead of traditional ">.50"

for(j in 1:iterations)
{
  trainIndices = sample(1:nrow(df1A), round(splitPerc * nrow(df1A)))
  train = df1A[trainIndices,]
  test  = df1A[-trainIndices,]
  
  model = naiveBayes(train[,3:17], as.factor(train$Attrition), laplace = 1)
  
  # Get predicted probabilities
  probs = predict(model, test[,3:17], type = "raw")
  
  # Apply custom threshold
  pred = ifelse(probs[, "Yes"] > threshold, "Yes", "No")
  pred = as.factor(pred)
  
  CM = confusionMatrix(pred, as.factor(test$Attrition), positive = "Yes")
  
  masterAcc[j]  = CM$overall["Accuracy"]
  masterSens[j] = CM$byClass["Sensitivity"]
  masterSpec[j] = CM$byClass["Specificity"]
  
  # Extract confusion matrix values
  masterTP[j] = CM$table["Yes","Yes"]
  masterTN[j] = CM$table["No","No"]
  masterFP[j] = CM$table["Yes","No"]
  masterFN[j] = CM$table["No","Yes"]
}

MeanAcc  = mean(masterAcc)
MeanSens = mean(masterSens)
MeanSpec = mean(masterSpec)

MeanAcc # mean acc
MeanSens # mean sens
MeanSpec # mean spec


MeanTP = mean(masterTP)
MeanTN = mean(masterTN)
MeanFP = mean(masterFP)
MeanFN = mean(masterFN)

AvgCM = matrix(c(MeanTP, MeanFP,
                 MeanFN, MeanTN),
               nrow = 2,
               byrow = TRUE)

rownames(AvgCM) = c("Pred Yes", "Pred No")
colnames(AvgCM) = c("Actual Yes", "Actual No")

AvgCM
```

### Creating Visuals for Slides
```{r}
# Changing Data to Long Format
library(tidyverse)

results_df <- data.frame(
  Accuracy = masterAcc,
  Sensitivity = masterSens,
  Specificity = masterSpec
)
# Option A: Histograms + Mean Line
results_long <- results_df %>%
  pivot_longer(cols = everything(),
               names_to = "Metric",
               values_to = "Value")

results_long$Value <- results_long$Value * 100

ggplot(results_long, aes(x = Value)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(Value)),
             color = "red",
             linewidth = 1) +
  facet_wrap(~Metric, scales = "free") +
  labs(title = "Distribution of Model Performance (100 Random Splits)",
       x = "Percentages",
       y = "Count") +
  theme_minimal(base_size = 14)
```


### Data Cleanup for KNN Model
```{r}
# Had to switch categorical variables to numeric for knn model
df1B <- df1A %>%
  mutate(
    Department = as.numeric(ordered(Department,
                          levels = c("Human Resources", "Research & Development", "Sales"))),
    EducationField = as.numeric(ordered(EducationField,
                             levels = c ("Human Resources", "Life Sciences", "Marketing", "Medical", "Other", "Technical Degree"))),
    OverTime = as.numeric(ordered(OverTime, 
                                  levels = c("No", "Yes"))),
    MaritalStatus = as.numeric(ordered(MaritalStatus,
                                       levels = c("Divorced", "Single", "Married"))),
    JobRole = as.numeric(ordered(JobRole,
                                 levels = c("Sales Executive", "Research Director", "Manufacturing Director", "Research Scientist", "Sales Representative", "Healthcare Representative", "Manager", "Human Resources", "Laboratory Technician")))
)
```

### KNN - Looking to see whcih value of K is best - df1
```{r}
library(class)
library(caret)

set.seed(6)

iterations  = 100

splitPerc = .7
threshold = 0.15   # custom decision threshold
numks = 20

masterAcc = matrix(nrow = iterations, ncol = numks)
masterSens = matrix(nrow = iterations, ncol = numks)
masterSpec = matrix(nrow = iterations, ncol = numks)

for(j in 1:iterations)
{
  trainIndices = sample(1:nrow(df1B), round(splitPerc * nrow(df1B)))
  train = df1B[trainIndices,]
  test  = df1B[-trainIndices,]
  
  # Standardize predictors (for KNN)
  trainX = scale(train[,3:12])
  testX = scale(test[,3:12])
  
  trainY = as.factor(train$Attrition)
  testY  = as.factor(test$Attrition)
  
  for (i in 1:numks)
  {
  # Run KNN
  knn_pred = knn(train=trainX, test=testX, cl=trainY, k=i, prob=TRUE)
  
  # Extract probability of winning class
  knn_prob = attr(knn_pred, "prob")
  
  # Convert to probability of "Yes"
  prob_yes = ifelse(knn_pred == "Yes",knn_prob, 1 - knn_prob)
  
  # Apply custom threshold
  pred = ifelse(prob_yes > threshold, "Yes", "No")
  pred = as.factor(pred)
  
  CM = confusionMatrix(pred, testY, positive = "Yes")
  
  masterAcc[j, i]  = CM$overall["Accuracy"]
  masterSens[j, i] = CM$byClass["Sensitivity"]
  masterSpec[j, i] = CM$byClass["Specificity"]
  }
}

MeanAcc  = colMeans(masterAcc)
MeanSens = colMeans(masterSens)
MeanSpec = colMeans(masterSpec)

plot(seq(1,numks,1),MeanAcc, type = "l", xlab = "k", main = "Mean Accuracy for K = 1-20")

plot(seq(1,numks,1),MeanSens, type = "l", xlab = "k", main = "Mean Sensitivity for K = 1-20")

plot(seq(1,numks,1),MeanSpec, type = "l", xlab = "k", main = "Mean Specificity for K = 1-20")

# lower k's look like better choice
```

### KNN = 4 - looking at  Accuracy, Sensitivity, Specificty - df1
```{r}
library(class)
library(caret)

set.seed(6)

iterations  = 100
# getting averages for acc, sens, spec
masterAcc   = numeric(iterations)
masterSens  = numeric(iterations)
masterSpec  = numeric(iterations)

# getting averages for confusion matrix
masterTP = numeric(iterations)
masterTN = numeric(iterations)
masterFP = numeric(iterations)
masterFN = numeric(iterations)


splitPerc = .7
threshold = 0.15   # custom decision threshold
k = 4              # choose your k

for(j in 1:iterations)
{
  trainIndices = sample(1:nrow(df1B), round(splitPerc * nrow(df1B)))
  train = df1B[trainIndices,]
  test  = df1B[-trainIndices,]
  
  # Standardize predictors (for KNN)
  trainX = scale(train[,3:12])
  testX = scale(test[,3:12])
  
  trainY = as.factor(train$Attrition)
  testY  = as.factor(test$Attrition)
  
  
  # Run KNN
  knn_pred = knn(train=trainX, test=testX, cl=trainY, k=k, prob=TRUE)
  
  # Extract probability of winning class
  knn_prob = attr(knn_pred, "prob")
  
  # Convert to probability of "Yes"
  prob_yes = ifelse(knn_pred == "Yes",knn_prob, 1 - knn_prob)
  
  # Apply custom threshold
  pred = ifelse(prob_yes > threshold, "Yes", "No")
  pred = as.factor(pred)
  
  CM = confusionMatrix(pred, testY, positive = "Yes")
  
  masterAcc[j]  = CM$overall["Accuracy"]
  masterSens[j] = CM$byClass["Sensitivity"]
  masterSpec[j] = CM$byClass["Specificity"]
  
  # Extract confusion matrix values
  masterTP[j] = CM$table["Yes","Yes"]
  masterTN[j] = CM$table["No","No"]
  masterFP[j] = CM$table["Yes","No"]
  masterFN[j] = CM$table["No","Yes"]
}

MeanAcc  = mean(masterAcc)
MeanSens = mean(masterSens)
MeanSpec = mean(masterSpec)

MeanTP = mean(masterTP)
MeanTN = mean(masterTN)
MeanFP = mean(masterFP)
MeanFN = mean(masterFN)

AvgCM = matrix(c(MeanTP, MeanFP,
                 MeanFN, MeanTN),
               nrow = 2,
               byrow = TRUE)

rownames(AvgCM) = c("Pred Yes", "Pred No")
colnames(AvgCM) = c("Actual Yes", "Actual No")

AvgCM

```


### Creating Visuals for Slides
```{r}
# Changing Data to Long Format

library(tidyverse)

results_df <- data.frame(
  Accuracy = masterAcc,
  Sensitivity = masterSens,
  Specificity = masterSpec
)

# Option A: Histograms + Mean Line

results_long <- results_df %>%
  pivot_longer(cols = everything(),
               names_to = "Metric",
               values_to = "Value")

results_long$Value <- results_long$Value * 100

ggplot(results_long, aes(x = Value)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(Value)),
             color = "red",
             linewidth = 1) +
  facet_wrap(~Metric, scales = "free") +
  labs(title = "Distribution of Model Performance (100 Random Splits)",
       x = "Percentages",
       y = "Count") +
  theme_minimal(base_size = 14)
```


## Competition Portion
```{r}
# Read in Data
df300 = read.csv('/Users/joshtsamuel/SMU/DS-6306 Doing Data Science/Units/Unit 8 & 9 - Case Study/CaseStudy1CompSet No Attrition.csv')

# Wrangling
df300 <- df300 %>%
  mutate(
    BusinessTravel = ordered(BusinessTravel,
                        levels = c("Non-Travel", "Travel_Frequently", "Travel_Rarely")),
    Department = ordered(Department,
                          levels = c("Human Resources", "Research & Development", "Sales")),
    EducationField = ordered(EducationField,
                             levels = c ("Human Resources", "Life Sciences", "Marketing", "Medical", "Other", "Technical Degree")),
  )

# Choosing same predictors 
df300 = df300 %>% select(ID, MonthlyIncome, TotalWorkingYears, OverTime, JobLevel, YearsAtCompany, YearsInCurrentRole, NumCompaniesWorked, EducationField, StockOptionLevel, Age, MaritalStatus, Department, JobRole, JobInvolvement, JobSatisfaction)


# Run Model
set.seed(6)
threshold = 0.30 
model = naiveBayes(df1A[,3:17], as.factor(df1A$Attrition), laplace = 1)

probs = predict(model, df300[,2:16], type = "raw")
pred = ifelse(probs[, "Yes"] > threshold, "Yes", "No")

results <- cbind(df300[, "ID", drop = FALSE], PredProbYes = probs,Attrition = pred)

# removing PredProb columns
results = results[, c(1,4)]

# writing to csv file
write.csv(results,
          '/Users/joshtsamuel/SMU/DS-6306 Doing Data Science/Units/Unit 8 & 9 - Case Study/Case1PredictionsSamuel Attrition.csv',
          row.names = FALSE)
```